---
title: Assesment-3 Part-1; importing files, data wrangling, mathematical operations,
  plots and saving code on GitHub
Theam: cosmo
output:
  html_document:
    df_print: paged
Output: html_document
---

Author: Yasodha Jayawardena & Dhanusha Weerawardana
---


# Introduction
Rstudio is a programming language which is appliable for projects on statisticals computing and graphics. Furthermore, the software was initially started on the year 2009 and according to Rstudio, their mission is to "create free and open-source software for data science, scientific research, and technical communication" with the hope of enhance production quality. 

However, during Assessment 3 part A, bioinformatics knowladge will be improve due to many activities on Rstudio software, licuding importing files, data wrangling, mathematical operations,
  plots and saving code on GitHub.

# Question 1-Download the file and read into the R studio.
In order to download the file by the comand `download.file` and then to import data into the R and cread data according to a format, respectivly `header = TRUE` and `row.name = 1` commands can be applied. In addition, `head()` and `str()` commands can be used to check if the data is improt properly and also the stucture of the data.

```{r,chunck1,echo=TRUE}
#Read the file
download.file("https://raw.githubusercontent.com/markziemann/SLE712_files/master/bioinfo_asst3_part1_files/gene_expression.tsv",destfile="gene_expression.tsv")
x<-read.table("gene_expression.tsv")
head(x)
str(x)

x <- read.table("gene_expression.tsv", header = TRUE , stringsAsFactors = FALSE , row.names = 1, sep="\t")
head(x)
str(x)
```


# Question 2-Make a new column for mean of the other columns and present a table with first 6 values.
In this situation, `$` apply create a data farame and then `<-` operator used to assign a new vector of data to it. All in all the command should appear like `_$Mean<- rowMeans ()`. Finally `head()` can be used to obtain mean values of first 6 columns.

```{r,chunck2,echo=TRUE}
#Create new column for mean values
x$Mean<- rowMeans(x)
head(x)
```

# Question 3-List the 10 genes with the highest mean expression.
in this question data has to be sort by giving command `order()` and the number of data can be determine by `n= required amount`. Altogteher the commanand should be `head(ord<-X[order(-X$Mean),],n=10)` and X is the data farme.

```{r,chunck3,echo=TRUE}
#10 genes with the higher mean values
head(ord<-x[order(-x$Mean),],n=10)
```

# Question 4-Determine the number of genes with a mean <10.
In order to obtain the value of mean value above 10, `which()` fuction can be used. To illustrate, in a logical vector `which()` fuction able to return the posion of true element.Thereby, the B is new data set(mean value above 10) and the data frame is `rowMean` of X.

```{r,chunck4,echo=TRUE}
#Number of gene with the mean value above 10
which(B<-rowMeans(x)<10)
```

# Question 5-Make a histogram plot of the mean values in png format.
in this question, plot a histogram for mean values and saved as in png format. Therefore, first `hist` command can be used to creat a histogram and `x$Mean` is the data frame. Secondly, `xlab` and `ylab` functions can be used to leble the x and y axis and finally in order to make the histogram clear, colors for border and columns can be apply.

To save the hitogram in to the R database, `png(location)` function can be used as the command for the formaat

```{r,chunck5,echo=TRUE}
#Histogram plot of mean value in png format
hist(x$Mean,breaks = 20,xlab="Genes",ylab="Mean",main="Mean values of the RNA sequency",col="green",border = "brown")
png(file="Assessment part 1 first section histogram.png")
hist(x$Mean,breaks = 20,xlab="Genes",ylab="Mean",main="Mean values of the RNA sequency",col="green",border = "brown")
dev.off()
```